The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-18 manuscript/manuscript.Rnw
add discussion of primate vs non-primate changes

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodona_africana),((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus)))),(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus),((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)))));")
plot(species.tree,x.lim=18)
edgelabels("Eutheria",2,frame="none",bg="white",font=1,adj=c(0.75,0),cex=0.9)
edgelabels("Euarchontoglires",16,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Laurasiatheria",7,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Atlantogenata",3,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Boreoeutheria",6,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

\subsection{Human Placenta Transcriptome}

>>>>

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodona_africana),((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus)))),(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus),((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)))));")
plot(species.tree,x.lim=18)
edgelabels("Eutheria",2,frame="none",bg="white",font=1,adj=c(0.75,0),cex=0.9)
edgelabels("Euarchontoglires",16,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Laurasiatheria",7,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Atlantogenata",3,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Boreoeutheria",6,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
edgelabels("Primates",22,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
edgelabels("Rodentia",17,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific")))>>=
load("../data/placenta_lineage_specific")
source("/home/don/projects/xtable/pkg/R/print.xtable.R")

placenta.lineage.specific.table <- function(analysis,output=TRUE) {
    temp.table <- 
        head(placenta.lineage.specific.analysis[analysis.name==analysis &
                                                p.value <= 0.05][order(p.value),
                                                                 list(human_name,p.value,fdr,fc),],
             n=5)
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setkey(temp.table,"Approved Symbol")
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    if (output) {
        print(xtable(temp.table[order(p.value),],
                     display=c("s","s","s","e","g","f"),
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

    }
    return(invisible(temp.table[order(p.value),]))
}
@

\subsection{Primate Placenta Transcriptomes}

\begin{table}
  \tiny
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- NULL
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
@ 
\caption{Genes whose expression changed significantly in the primate
  order as compared to all other analyzed species.}
\label{tab:primate_lineage_specific}
\end{table}

We examined the expression of three primates, \human, \bonobo, and
\smonkey. The genes whose expressions are different on the primate
lineage in comparison to all other analyzed species are listed in
\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
higher expression in the primate lineage
(\cref{fig:fstl1_evolution_tree},
$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1","FDR"]}$), is implicated
in BMP signaling, and has previously been implicated in inflammation
and invasiveness
\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
non-primates, \horse shows the highest expression
%
\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
%
which is still lower than the lowest primate, \smonkey
%
(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}. 
%
No other genes were significantly differentially expressed on this
lineage by t-test.

\begin{figure}
<<fstl1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="FSTL1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
@ 
\caption{Expression of follistatin-like 1 \textit{FSTLl1} and related
  genes.}
\label{fig:fstl1_evolution_tree}
\end{figure}

\subsubsection{\human Placenta Transcriptome}
\label{sec:human_placenta}
